TAPIR: Threaded Application Programming on Intermittent Resources

### 2017-05-11 paper.tex
updated references and related work

\section{Introduction/Background}
  \label{sec:intro}
%A/N: I don't think that the paragraphs in this section are in a great order, but I am having a hard time making it flow better.
%Also I'm absolutely sure each of these paragraphs can definitely be bulked up more for length. Please help thank you!!!
%Also also: I'm trying to get [energy-harvesting computing devices] and [intermittent execution model] below to be in italics,
%but when I use \textit it makes the ENTIRE PAPER in italics, i used the curly braces and everything helpppppp that has never happened to me before

The increase in popularity of Internet of Things (IoT) has led to research focus in wireless technologies and low-power computing.
New technologies in both of these areas has led to new [energy-harvesting computing devices] that operates only with power it has 
collected from the surrounding environment \cite{Chain}. These devices have use cases in medical, IoT, and research applications.

Commonly, an energy-harvesting devices charges and collects power in a storage capacitor until the capacitor reaches a sufficient level of charge.
Once the capacitor is charged, the device begins its operation phase and will run until the capacitor is drained.
This leads to the \textit{intermittent execution model} where regular shut downs and reboots to recharge are the common case rather than the rare case in
a traditional execution model. Because the intermittent execution model includes power failures and reboots, continuous execution programs behave
irregular behavior on an intermittent device and a new programming model is needed.

Our multithreaded model is build on two existing programming models, DINO and Chain.
An initial intermittent programming model DINO breaks intermittent programs into tasks, which creates automatic checkpointing locations 
during exeuction \cite{Dino}. DINO uses a checkpointing and recovery model that tracks both volatile and nonvolatile memory.
The DINO execution model executes all tasks transactionally, leading to atomic task semantics and does not permit consistency violations in 
nonvolatile state \cite{Dino}. Chain, a more recent iteration of the concepts in DINO, uses volatile state rarely, if ever, and uses task boundaries
called channels to store most used data in nonvolatile memory \cite{Chain}. At compile time, Chain allocates the necessary channels between tasks
by parsing channel read and write statements in order to match up aligning tasks. All inputs and outputs to each task, as well as necessary data
structures, are stored in a channel for a respective task. A channel may only be modified by one task, and only one task may run at a time,
which mitigates time of check, time of use bugs (TOCTOU) in intermittence. 
The one exception is a self-channel for recursive code, which is double buffered.
Keeping all modifiable data in nonvolatile memory that is not open to easy rewrite makes these tasks idempotent across reboots.
Because of continuous reboots, many operations necessarily will be executed more than once,
and correctness is compromised if those operations are not guaranteed to have the same outcome by the programming model after each reboot.

A main issue in intermittent computing along with idempotence is the Read-Modify-Write (RMW) problem. 
Because volatile memory is cleared on each reboot, it is possible that in between reading and modifying or modifying and writing the data had changed,
leading to incorrect answers and an inconsistent state. In the Chain programming model, this problem is solved by creating a new task to house the
problematic operations. Because tasks are blocked by channels that can only be written by one task and are saved to nonvolatile memory, RMW is avoided.
However, in multiprogramming the problem manifests in another way; if the scheduler switches tasks on a reboot, the currently running thread could have
been in the middle of its critical section, making it difficult for other threads to make progress if context switches always occur after a reboot.
Because of this we have decided on a round-robin scheduler with context switches at the end of a task, to guarantee forward progress while knowing
which tasks have run to completion.

In the current Chain programming model, writing code that completes multiple objectives correctly is now difficult. 
Current chain programs are small and only have one objective. Additionally, there is no support for advanced control flow.
Multiple control threads may exist, but in the current model there is no support for them and programmability is difficult and often incorrect.
Supporting multiple control threads in Chain also necessitates a linear increase in a limited memory resource, because all large shared data structures
must be copied into each necessary channel, and also makes updating the structures a time consuming, high-overhead task.
In order to increase functionality of chain and to mitigate these shortcomings, we add familiar threading semantics to Chain in order to express
multi-threaded control flow operation. In order to do this we create a lightweight scheduler task that becomes part of the inter-task checkpointing
operation and provide a method of sharing large data structures between tasks with a locking feature. This method is low-memory and low-execution
overhead, which can then maintain the Chain programming semantics at the user level still successfully complete work on an intermittent device.

Example Citation\cite{Chain, Mementos, Aware, Dino}!

@@

\section{Related Work} -Emily
Operating system work\\
TU delft guys...\\
Energy Align\\

@@

\section{What we did} % TODO change title
Scheduler\\
Mutexes - Emily\\
User programming model for threads - Neil\\
Round Robin (alternatives, on reboot)\\
SIMD vs MIMD\\
SIMD w/ channels (you don't get your own registers...etc -
    just channels, since intermittent)\\

\subsection{Interrupts}
\subsubsection{Motivation}
Ideally, multithreading would extend to a programming model that can utilize
interrupts in a producer/consumer fashion (i.e. a thread is waiting on data
from an external source, the external source fires an interrupt, the thread
samples the external source and stores the received data in a buffer, and
the thread returns to its original task). However, the inherent
non-deterministic
nature of interrupts from, say, an ADC, creates additional complexity for the
programmer when reasoning about shared memory, especially with respect to
idempotence.\\
    Specific to the MSP430 architecture (but common in other systems), there is
also the issue of stack integrity. On the MSP430, when an interrupt fires,
the program counter and status register are pushed onto the stack for
the handler to be able to return back to the state of the machine when the
interrupt fired. In our case, however, where power failure is the common
case, an application could experience failure while inside an interrupt
handler. This raises two issues. First, upon reboot from a power failure,
the stack will be cleared (since the stack is stored in volatile memory) and
execution will return to the handler, if we assume that the interrupt handler
is a \chain task. Upon leaving the interrupt, via the MSP430 RETI instruction),
the program's behavior will be undefined. Second, if the programmer does not
use a \chain task for the interrupt handler, it is possible that the interrupt
could fire when the system has a low amount of energy remaining, thus
rendering the system unable to finish executing the interrupt handler,
leading to a host of potential problems.

>>>>

\section{Introduction/Background} -Amanda
  \label{sec:intro}
  Chain, DINO\\
  Multithreading\\
  Intermittance Review\\
  RMW\\
  Example Citation\cite{Chain, Mementos, Aware, Dino}!

@@

\section{Related Work} -Emily
In addition to the division of a program into tasks, as described above, prior work
explores several different strategies for preserving state across reboots. The approaches
range from fully independent of the programmer~\cite{ratchet, dewdrop} at one extreme, to programmer
defined task boundaries~\cite{dino} at the other.  Compiler driven approaches such as the
method used in ~\cite{Mementos} allow the programmer to specify high level behavior of the
checkpointing system. A key similarity among the prior work is it
does not lend itself well to multi-programming let
alone multi-threading. In fact, many of these works state the assumption that the device
hardware cannot support a conventional OS, and thus focus on single threaded programs. 
Despite the availability of (relatively) inexpensive nonvolatile memory technology such as
FRAM~\cite{quickrecall}, the the energy and storage overhead of checkpointing multiple
thread states would limit the maximum per-thread checkpoint size.

Furthermore, if a multiprogrammed application were implemented using previous
checkpointing strategies, reasoning about thread interleavings becomes  difficult to
reason about because of unpredictable power failures. Any conventional timeout based
thread switching approach will be challenging in an intermittent system, but systems that
allows dynamic checkpointing~\cite{hibernus}, or dynamic task formation~\cite{Dino} become
even more difficult to program. Particularly in systems like ~\cite{Dino} where the
programmer is responsible for placing task boundaries, there is a risk that programmers
will either overuse task boundaries in an effort to preserve any given thread's state, or
underprovision task boundaries such that progress is never made because of context
switching that dynamically expands the number of instructions between task boundaries.

While prior work has stopped short of suggesting an operating system for intermittently
powered devices, a number of different works touch on the need for these devices to
perform multiple tasks in the same workload. Buettner, Greenstein and Wetherall motivate
their runtime, Dewdrop, based on difference in energy requirements for the tasks that a
CRFID needs to perform~\cite{dewdrop}. The energy aware memory mapping in ~\cite{Aware}
discusses the challenge of handling nondeterministic interrupts received by intermittent
sensing platforms when a peripheral must be serviced immediately and control flow deviates
from single threaded execution. From a hardware standpoint, the federated energy scheme in
~\cite{ufop} provides a platform for removing the energy dependence between an MCU and its
peripherals. Federated energy also simplifies a programmer's reasoning about reading from
a single peripheral. However, federating energy does not help a programmer describe how to
handle {\em multiple} peripherals. Consider a listing in
~\cite{ufop} reporoduced in Figure ~\ref{copy}, and imagine the level of understanding
of the hardware behavior a programmer would have to employ to, say, read one sensor if it
is available but read from and proceed to process data from a different sensor if it is not.

Operating system work\\
Want to do multiple things- i.e. federated energy \\
Dewdrop- energy aware task scheduling\\
TU delft guys...\\
Energy Align\\
TinyOS stuff in wireless sensor networks\\

@@

\section{What we did} % TODO change title
Scheduler\\
Mutexes - Emily\\
User programming model for threads - Neil\\
Round Robin (alternatives, on reboot)\\
SIMD vs MIMD\\
SIMD w/ channels (you don't get your registers...etc -
    just channels, since intermittent\\
\subsection{Interrupts}
\subsubsection{Motivation}
Ideally, multithreading would extend to a programming model that can utilize
interrupts in a producer/consumer fashion (i.e. a thread is waiting on data
from an external source, the external source fires an interrupt, the thread
samples the external source and stores the received data in a buffer, and
the thread returns to its original task). However, the inherent
non-deterministic
nature of interrupts from, say, an ADC, creates additional complexity for the
programmer when reasoning about shared memory, especially with respect to
idempotence.\\
    Specific to the MSP430 architecture (but common in other systems), there is
also the issue of stack integrity. On the MSP430, when an interrupt fires,
the program counter and status register are pushed onto the stack for
the handler to be able to return back to the state of the machine when the
interrupt fired. In our case, however, where power failure is the common
case, an application could experience failure while inside an interrupt
handler. This raises two issues. First, upon reboot from a power failure,
the stack will be cleared (since the stack is stored in volatile memory) and
execution will return to the handler, if we assume that the interrupt handler
is a \chain task. Upon leaving the interrupt, via the MSP430 RETI instruction),
the program's behavior will be undefined. Second, if the programmer does not
use a \chain task for the interrupt handler, it is possible that the interrupt
could fire when the system has a low amount of energy remaining, thus
rendering the system unable to finish executing the interrupt handler,
leading to a host of potential problems.
